---
title: Practical Machine Learning Project
output: html_document
---
In this project, our goal is to use exercise data from accelerometers on the belt, forearm, arm, and dumbell of 6 participantsThe goal of this project is to predict the manner in which they did the exercise. 

This report aim to describe,
How the model is built, 
how did I use cross validation, 
Thoughts about expected out of sample error 
why I made the choices we did. 
I also did use our prediction model to predict 20 different test cases.

Reproduceability

An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below

Packages such as caret and randomForest were installed

Model

The variable classe, a factor variable with 5 levels. For this data set, "participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes." [1]
Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as my final model.


Cross-validation is performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

Expected out-of-sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

Reasons for my choices

Our outcome variable "classe" is an unordered factor variable. Thus, we can choose our error type as 1-accuracy. We have a large sample size with N= 19622 in the Training data set. This allow us to divide our Training sample into subTraining and subTesting to allow cross-validation. Features with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables.
Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification [2]. Feature selection is inherent, so it is not so necessary at the data preparation phase. Thus, there won't be any feature selection section in this report.

Input Data
```{r}
library(caret)
library(randomForest)
```
```{r}
library(rpart)
library(rpart.plot)
```
```{r}
set.seed(1234)
```

Reading, Loading and Cleaning data
```{r}
trainingset <- read.csv("C://Users//Indika//Desktop//Coursera//Practical Machine learning Project//pml-training.csv", na.strings=c("NA","#DIV/0!", ""))
```
```{r}
testingset <- read.csv("C://Users//Indika//Desktop//Coursera//Practical Machine learning Project//pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))
```

```{r}
dim(trainingset)
dim(testingset)
```
```{r}
trainingset<-trainingset[,colSums(is.na(trainingset)) == 0]
testingset <-testingset[,colSums(is.na(testingset)) == 0]
```
```{r}
trainingset   <-trainingset[,-c(1:7)]
testingset <-testingset[,-c(1:7)]
```
```{r}
dim(trainingset)
dim(testingset)
head(trainingset)
head(testingset)
```
Partition for Cross validation (training Data)
```{r}
subsamples <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
subTraining <- trainingset[subsamples, ] 
subTesting <- trainingset[-subsamples, ]
dim(subTraining)
dim(subTesting)
head(subTraining)
head(subTesting)
```

Data Plot
```{r}
plot(subTraining$classe, col="green", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency")
```

Prediction Model 1 (Decison tree)
```{r}
model1 <- rpart(classe ~ ., data=subTraining, method="class")
```
```{r}
prediction1 <- predict(model1, subTesting, type = "class")
```
```{r}
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

Test Results form Subsetting data
```{r}
confusionMatrix(prediction1, subTesting$classe)
```

Prediction Model 2-Random forest
```{r}
model2 <- randomForest(classe ~. , data=subTraining, method="class")
```
```{r}
prediction2 <- predict(model2, subTesting, type = "class")
```
```{r}
confusionMatrix(prediction2, subTesting$classe)
```

Decision

As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.739 (95% CI: (0.727, 0.752)) for Decision Tree model. The random Forest model is choosen. The accuracy of the model is 0.995. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.



```{r}
predictfinal <- predict(model2, testingset, type="class")
predictfinal
```

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)
```

